---
title: "Translating brown bear genotypes"
author: "Roman Luštrik & Marta de Barba"
date: "2020-07-14"
output: html_document
---

```{r setup, include=FALSE}
library(readxl)
library(tidyr)
library(transgt)

knitr::opts_chunk$set(echo = TRUE)
```

This script translates brown bear genotypes. Data comes from U Grenoble database and uses a manually constructed reference of loci translations.

# Import and prepare data
```{r}
microsat.binary <- "./grenoble_microsat.RData"

if (!file.exists(microsat.binary)) {
  source("import_from_dataset.R")
} else {
  load(microsat.binary)
  # Populates global environment with `microsat` object.
}

# Fetch translation_table tab of
# https://docs.google.com/spreadsheets/d/18wZNXmmYMYB1AmVCisQMLHg-swj9HswLUXjMz0akhVI/edit#gid=0
# and save it as (date)_reference_table.xlsx file.
reftbl <- read_excel(path = "2020_06_27_reference_table.xlsx", sheet = "translation_table")
reftbl <- as.data.frame(reftbl)

# Table needed to translate population names to names used in reference table.
maptbl <- read_excel(path = "2020_06_27_reference_table.xlsx", sheet = "grenoble_dataset_lab_names")
maptbl$short_name <- NULL

# Genotype data
head(microsat)
microsat$loci_name <- trimws(microsat$loci_name)
microsat$research_group <- trimws(microsat$research_group)
```

# Prepare reference table
In order to translate microsatellite data, we also need population name that matches that in the reference table. Translation will happen only for data that have mapping in the reference table.
```{r}
ms <- merge(x = microsat, y = maptbl, by.x = "research_group", by.y = "database_name", sort = FALSE, all = TRUE)
ms <- ms[!is.na(ms$lab_from), ]
```

```{r}
# This is because we need allele data in LOC_1 LOC_2 format where each entry is
# its own column.
xy <- pivot_wider(
  data = ms,
  id_cols = c(lab_from, id_sample),
  names_from = loci_name,
  values_from = c(size_all1, size_all2))

xy <- as.data.frame(xy)

colnames(xy) <- gsub(pattern = "^(size_all)(\\d)_(.*)$",
     replacement = "\\3_\\2",
     x = colnames(xy))

xy.cols <- colnames(xy)

colnames(xy)[colnames(xy) == "id_sample"] <- "sample"

# out <- translateGenotypes(input = xy, ref_tbl = reftbl, long = TRUE)
out <- translateGenotypes(input = xy, ref_tbl = reftbl, long = FALSE)

# Prepare data for output
out$allele <- NULL
out$side <- gsub("^.*(_\\d)$", replacement = "\\1", x = out$locus)
out$locus <- gsub("^(.*)_\\d$", replacement = "\\1", x = out$locus)

out <- pivot_wider(data = out, 
              values_from = translated_allele,
              names_from = side
  )

colnames(out)[colnames(out) == "_1"] <- "trans_size_all1"
colnames(out)[colnames(out) == "_2"] <- "trans_size_all2"
colnames(out)[colnames(out) == "locus"] <- "loci_name"
colnames(out)[colnames(out) == "sample"] <- "id_sample"

out <- merge(x = out, y = ms,
      by = c("lab_from", "id_sample", "loci_name"),
      sort = FALSE,
      all = TRUE)

out <- out[, c(colnames(ms), "trans_size_all1", "trans_size_all2")]
out$lab_from <- NULL

save(out, file = "2020_06_28_translated_genotypes.RData")

write.table(x = out, file = "2020_06_28_translated_genotypes.txt", row.names = FALSE, quote = FALSE, sep = "\t")
```

```{r}
microsat[microsat$research_group == "University of Ljubljana · Department of Biology", "id_sample"]
microsat[microsat$id_sample == 721, ]
```

